---
title: "Module 3: Data-Limited Stock Assessment Methods"
subtitle: "Somali Fish Stock Training"
date: "2025-12-05"
date-modified: today
author: 
    - name: "Dr. Masumbuko Semba"
      affiliation: "Nelson Mandela African Institution of Science and Technology"
      email: "lugosemba@gmail.com"
    - name: "Dr. Rushingisha George"
      affiliation: "Tangania Fisheries Research Institute"
      email: "george.rushingisha@tafiri.go.tz"
    - name: "Dr. Mary Kishe"
      affiliation: "Tanzania Fisheries Research Institute"
      email: "mary.kishe@tafiri.go.tz"
    - name: "Dr. Ismael Kimirei"
      affiliation: "Tanzania Fisheries Research Institute"
      email: "ismael.kimirei@tafiri.go.tz"
format: revealjs
theme: default
transition: slide
background-transition: fade
highlight-style: github
embed-resources: true
---

```{r}
require(sf)
require(tidyverse)
require(sizeMat)
require(gt)
require(tidyplots)
require(patchwork)
require(leaflet)
```

## Welcome to Module 3

**Data-Limited Stock Assessment Methods**

:::: {.columns}
::: {.column width="50%"}
### Learning Objectives
- Apply length-based assessment models
- Use empirical indicators for status
- Conduct practical R-based analyses
- Make evidence-based recommendations
:::

::: {.column width="50%"}
### Why This Matters
- **Most requested** training component
- **Lowest expertise** area identified
- **Directly applicable** to Somali data
- **Actionable** for decision-makers
:::
::::

---

## What Are Data-Limited Methods?

Stock assessments when you have:

::: {.columns}
::: {.column width="50%"}
**What You DON'T Have:**
- ✗ Age structure data
- ✗ Long time series
- ✗ Detailed catch history
- ✗ Tagging studies
- ✗ Population surveys
:::

::: {.column width="50%"}
**What You DO Have:**
- ✓ Length measurements
- ✓ Catch data
- ✓ Basic life history
- ✓ Landing information
- ✓ Recent time series
:::
::::

---

## Why Length-Based Methods?

### Length is Easy to Measure:

::: {.incremental}
- Simple equipment (measuring ruler)
- Minimal training required
- Immediate feedback
- Low cost
- High consistency
:::

### Length Contains Information:

Size distributions reveal:
- Population age structure
- Growth rates
- Fishing pressure
- Recruitment patterns

---

## Key Concept: Growth

### Von Bertalanffy Growth Model:

Fish length increases over time following a curve:

$$L_t = L_∞ (1 - e^{-K(t-t_0)})$$

Where:
- **L∞**: Asymptotic length (max size)
- **K**: Growth rate (how fast)
- **t**: Age
- **t₀**: Age at length zero

---

## Length-Based Methods Overview

### Three Main Approaches:

::: {.columns}
::: {.column width="33%"}
**1. LBB Model**
Length-Based Bayesian
- Estimates from data
- Complex but flexible
- Best when trained
:::

::: {.column width="33%"}
**2. LIME Model**
Length-Integrated Mixed Effects
- Uses length frequencies
- Integrated framework
- Well-developed
:::

::: {.column width="33%"}
**3. Empirical Indicators**
Simple heuristics
- Easy to calculate
- Transparent
- Immediate insights
:::
::::

---

## Empirical Indicators: Length Metrics

### Mean Length:

Average size of fish in catch

```r
mean_length <- data |>
  group_by(year) |>
  summarise(mean_L = mean(length_cm, na.rm = TRUE))
```

**Interpretation:**
- Increasing trend = good (growing population)
- Decreasing trend = concern (overfishing)

---

## Empirical Indicators: Median Length

### Median Length:

Middle value of size distribution

:::: {.columns}
::: {.column width="50%"}
**Robust to outliers**
- Not affected by very large fish
- Better than mean for skewed data
- Easier to communicate
:::

::: {.column width="50%"}
**Calculate in R:**
```r
median_length <- data |>
  group_by(year) |>
  summarise(
    med_L = median(length_cm, na.rm = TRUE)
  )
```
:::
::::

---

## Empirical Indicators: L95 & Lmax

### L95 (95th Percentile):

Size at which 95% of fish are smaller

::: {.incremental}
- Indicates presence of large fish
- Decreasing L95 = concerning pattern
- Vulnerable to sampling gaps
:::

### Lmax (Maximum Observed):

Largest fish in catch

- Changes with sampling intensity
- Indicator of recruitment & growth
- Use cautiously (influenced by luck)

---

## Empirical Indicators: Proportion Mature

### Size-at-Maturity Concept:

Many fish below breeding size = recruitment problem

```r
# Define maturity based on size threshold
data_mature <- data |>
  mutate(
    mature = if_else(length_cm >= 35, "Mature", "Immature")
  ) |>
  group_by(year, mature) |>
  summarise(count = n(), .groups = 'drop') |>
  pivot_wider(names_from = mature, values_from = count) |>
  mutate(prop_mature = Mature / (Mature + Immature))
```

---

## Empirical Indicators: Catch Composition

### Species Dominance:

Track which species dominate catch

```r
composition <- data |>
  filter(!is.na(fish_group)) |>
  group_by(year, fish_group) |>
  summarise(count = n(), .groups = 'drop') |>
  group_by(year) |>
  mutate(proportion = count / sum(count))
```

**Red Flags:**
- One species >50% = concentration risk
- Rapid shifts = ecosystem change
- Loss of diversity = vulnerability

---

## Practical Example: Somali Data Analysis

### Our Dataset Recap:

```{r}
#| echo: false
#| eval: false

# This represents your actual data
tibble(
  Metric = c("Records", "Years", "Regions", "Species", "Avg Length", "Avg Weight"),
  Value = c("3,571", "2017-2023", "6", "7 groups", "32.4 cm", "8.2 kg")
) |>
  gt() |>
  tab_header(title = "Somali Landing Data Summary") |>
  opt_stylize(style = 6, color = 'gray')
```

---

## Analysis 1: Mean Length Trends

### R Code Example:

```r
length_trends <- lw |>
  filter(!is.na(length_cm), !fish_group == 'Rays and Skates') |>
  group_by(year, fish_group) |>
  summarise(
    mean_length = mean(length_cm, na.rm = TRUE),
    n = n(),
    .groups = 'drop'
  )

# Visualize
length_trends |>
  tidyplot(x = year, y = mean_length, color = fish_group) |>
  add_data_points(size = 2) |>
  add_curve_fit(method = 'loess') |>
  adjust_title('Mean Length Trends')
```

---

## Analysis 2: Size Structure Changes

### Length-Frequency Distributions:

```r
lw |>
  filter(year %in% c(2017, 2023), !is.na(length_cm)) |>
  tidyplot(x = length_cm, fill = factor(year)) |>
  add_histogram(alpha = 0.6, binwidth = 5) |>
  adjust_title('Size Structure: 2017 vs 2023') |>
  adjust_x_axis(title = 'Length (cm)') |>
  adjust_y_axis(title = 'Frequency')
```

---

## Analysis 3: Maturity-Based Assessment

### Proportion of Fish Above Maturity Size:

```r
# Assume maturity at 30cm for demonstration
maturity_threshold <- 30

maturity_analysis <- lw |>
  filter(!is.na(length_cm)) |>
  mutate(mature = length_cm >= maturity_threshold) |>
  group_by(year) |>
  summarise(
    prop_mature = sum(mature) / n(),
    n_immature = sum(!mature),
    .groups = 'drop'
  )
```

**Interpretation:**
- Declining prop_mature = recruitment failure
- High immature % = good recruitment
- Threshold varies by species

---

## Analysis 4: Growth Estimation

### Length-Weight Relationships:

```r
# Fit allometric model: W = a * L^b
growth_model <- lw |>
  filter(!is.na(length_cm), !is.na(weight_kg)) |>
  lm(log(weight_kg) ~ log(length_cm), data = .)

# Extract parameters
a <- exp(coef(growth_model)[1])
b <- coef(growth_model)[2]

# Interpretation: b ~= 3 (isometric growth)
# b < 3 = getting thinner as grow (poor condition)
# b > 3 = getting heavier as grow (good condition)
```

---

## Analysis 5: Regional Comparison

### Stock Status by Region:

```r
regional_status <- lw |>
  filter(!is.na(region), !is.na(length_cm)) |>
  group_by(region) |>
  summarise(
    mean_length = mean(length_cm),
    median_length = median(length_cm),
    l95 = quantile(length_cm, 0.95),
    n = n(),
    cpue_2023 = sum(year == 2023) / n_distinct(landing_site),
    .groups = 'drop'
  )
```

---

## Traffic Light Assessment

### Simple Decision Framework:

```r
regional_status <- regional_status |>
  mutate(
    status = case_when(
      mean_length > 35 & cpue_2023 > 20 ~ "GREEN: Healthy",
      mean_length > 30 & cpue_2023 > 15 ~ "YELLOW: Caution",
      TRUE ~ "RED: Concerning"
    )
  )
```

---

## Interpreting Length-Based Results

### What Mean Length Tells Us:

::: {.columns}
::: {.column width="50%"}
**Increasing Mean Length**
- ✓ Stock recovering
- ✓ Good recruitment
- ✓ Reduced fishing
- ✓ Positive trend
:::

::: {.column width="50%"}
**Decreasing Mean Length**
- ✗ Fishing pressure high
- ✗ Recruitment issues
- ✗ Population aging
- ✗ Negative trend
:::
::::

---

## Reference Points Concept

### Management Targets:

Define desired stock conditions:

::: {.incremental}
- **Target reference point (TRP)**: What we aim for
- **Limit reference point (LRP)**: Below this = risky
- **Precautionary approach**: Stay well above LRP
:::

### Example from Data:
```r
# Hypothetical reference points
target_mean_length <- 35  # cm
limit_mean_length <- 25   # cm

status <- mean_length |>
  mutate(
    risk = case_when(
      mean_L > target_mean_length ~ "Low",
      mean_L > limit_mean_length ~ "Medium",
      TRUE ~ "High"
    )
  )
```

---

## Uncertainty in Assessments

### Sources of Uncertainty:

::: {.columns}
::: {.column width="50%"}
**Data Uncertainty**
- Sampling bias
- Measurement error
- Incomplete coverage
- Missing years
:::

::: {.column width="50%"}
**Model Uncertainty**
- Simplifying assumptions
- Parameter estimates
- Biological variability
- Environmental factors
:::
::::

### Best Practice:
Always communicate uncertainty with estimates!

---

## Decision-Making Under Uncertainty

### Precautionary Principle:

When uncertain, be conservative:

```r
# Don't just report point estimate
result <- tibble(
  estimate = 32.5,
  lower_ci = 30.2,
  upper_ci = 34.8,
  recommendation = "Maintain current effort, monitor closely"
)
```

---

## Practical Workshop: Your Assessment

### Exercise Data:

Use the Somali landing data to assess one species

**Steps:**
1. Filter to one fish group
2. Calculate key indicators
3. Interpret results
4. Recommend management action

### Deliverable:
One-page stock assessment report

---

## Stock Assessment Report Template

### Minimum Contents:

::: {.callout-note}
**1. Executive Summary** (1 paragraph)
- Stock status: Green/Yellow/Red
- Key findings
- Main recommendation

**2. Data & Methods** (brief)
- Data sources
- Indicators used
- Time period

**3. Results** (with figures)
- Key metrics & trends
- Comparisons with history

**4. Conclusions**
- Stock status interpretation
- Uncertainties
- Management recommendations

**5. References**
- Data sources
- Methods citations
:::

---

## Case Study: Example Report

### Berbera Tuna Stock:

::: {.callout-important}
**Status: YELLOW (Caution)**

Mean length declining 2.1 cm/year (2017-2023)
- 2017: 37.2 cm
- 2023: 34.5 cm

Interpretation: Increasing fishing pressure or recruitment declining

Recommendation: 
- Monitor closely next 6 months
- Reduce catch if trend continues
- Increase sampling frequency
:::

---

## Communicating Results to Decision-Makers

### Avoid Technical Jargon:

Instead of:
> "The von Bertalanffy growth model shows K=0.28, indicating asymptotic length of 55cm..."

Say:
> "Fish are growing to about 55cm maximum. They reach adult size at 5-6 years old."

---

## Visualizations for Policy Makers

### Keep It Simple:

::: {.incremental}
- **One main message per chart**
- **Use traffic light colors** (Red/Yellow/Green)
- **Show trends clearly**
- **Label axes in plain language**
- **Include uncertainty** (error bars)
- **Highlight key changes**
:::

---

## Management Actions Based on Assessment

### Green Status Actions:

```
✓ Maintain current fishing levels
✓ Continue regular monitoring
✓ Share success with community
✓ Document best practices
```

---

## Management Actions: Yellow Status

### Caution Response:

```
⚠ Reduce fishing effort by 10-20%
⚠ Increase monitoring frequency
⚠ Investigate cause of decline
⚠ Consult stakeholders
```

---

## Management Actions: Red Status

### Urgent Response:

```
✗ Implement immediate catch reductions
✗ Restrict fishing areas or seasons
✗ Emergency data collection
✗ Convene management committee
```

---

## Long-Term Monitoring Strategy

### Building Your Program:

**Year 1:** Establish baseline
- Collect data at all sites
- Train staff
- Test protocols

**Year 2-3:** Annual assessments
- Compare with baseline
- Detect trends
- Refine methods

**Year 4+:** Adaptive management
- Adjust regulations
- Scale sampling
- Expand to new areas

---

## R Resources for Module 3

### Useful Packages:

```r
# Length-based analysis
library(TropFishR)  # Von Bertalanffy, LBB
library(sizeMat)    # Maturity analysis

# Data visualization
library(ggplot2)    # Publication-quality plots
library(tidyplots)  # Simple, professional graphics

# Statistical summaries
library(rstatix)    # Quick statistics
library(gt)         # Professional tables
```

---

## Available Software Tools

### Free & Open Source:

- **LBSPR** (Length-Based Spawning Potential Ratio)
- **LBB** (Length-Based Bayesian)
- **SAMtool** (Stock Assessment Methods)
- **FishLife** (Life history estimation)

All work in R or standalone

---

## Key Takeaways

### Module 3 Summary:

::: {.incremental}
1. Data-limited methods work with what you have
2. Length-based indicators reveal stock status
3. Simple metrics often more useful than complex models
4. Uncertainty is inevitable—communicate it clearly
5. Trends matter more than absolute values
6. Regular monitoring enables adaptive management
7. Transparent methods build stakeholder confidence
:::

---

## Integrating All Three Modules

### Complete Assessment Framework:

:::: {.columns}
::: {.column width="33%"}
**Module 1**
Concepts
- Stock assessment principles
- Key indicators
- Data types
:::

::: {.column width="33%"}
**Module 2**
Operations
- Sampling design
- Data quality
- Field protocols
:::

::: {.column width="33%"}
**Module 3**
Analysis
- Length methods
- Indicators
- Management
:::
::::

---

## Building Capacity in Somalia

### Realistic Implementation:

**Immediate (0-3 months):**
- Train 1-2 lead analysts
- Set up R environment
- Conduct first assessments

**Short-term (3-12 months):**
- Expand to all regions
- Develop annual reports
- Build stakeholder engagement

**Long-term (1-2 years):**
- Integrate into policy
- Adaptive management cycle
- Regional coordination

---

## Next Steps for Participants

### After Training:

::: {.callout-note}
1. **Practice**: Analyze data from your region
2. **Collaborate**: Share findings with colleagues
3. **Document**: Create assessment procedures
4. **Communicate**: Present to decision-makers
5. **Iterate**: Refine methods based on feedback
6. **Scale**: Expand to other stocks
:::

---

## Resources & Continued Learning

### Online Training:
- FAO e-learning modules
- NOAA fisheries courses
- R programming resources

### Publications:
- Froese & Pauly (FishBase)
- Hordyk et al. (LBB methods)
- Costello et al. (Data-limited approaches)

### Networking:
- IOTC (Indian Ocean Tuna Commission)
- Regional fisheries bodies
- International assessment workshops

---

## Common Pitfalls to Avoid

### Don't:

::: {.incremental}
- Over-interpret single year changes
- Ignore measurement uncertainty
- Use methods beyond your understanding
- Skip quality control checks
- Extrapolate beyond data range
- Report false precision
- Ignore biological reality
:::

---

## Success Indicators

### How to Know It's Working:

- ✓ Decision-makers use assessments
- ✓ Fishers understand recommendations
- ✓ Data quality improving annually
- ✓ Stocks responding to management
- ✓ Stakeholders engaged
- ✓ Capacity growing locally
- ✓ Methods documented & standardized

---

## Troubleshooting Guide

### Problem: No Obvious Trend

**Cause:** High variability, short time series
**Solution:** Use longer window, increase sampling

### Problem: Conflicting Indicators

**Cause:** Different metrics tell different stories
**Solution:** Investigate causes, use weight of evidence

### Problem: Results Don't Match Expectations

**Cause:** Data issues, biological reality
**Solution:** Quality check, consult experts, revise understanding

---

## Questions for Your Assessment

### Before Making Recommendations:

- What changed and why?
- Is the trend significant or noise?
- What's the uncertainty?
- Have I verified the data?
- Do results match other evidence?
- Are there alternative explanations?
- What will decision-makers do with this?
- How confident am I in the conclusion?

---

## Course Wrap-Up

### You Have Learned:

- ✓ **Module 1**: Conceptual foundations
- ✓ **Module 2**: Operational protocols
- ✓ **Module 3**: Practical analysis methods

### You Can Now:

- Design monitoring programs
- Collect quality data
- Analyze fish populations
- Assess stock status
- Make recommendations
- Communicate findings

---

## Celebration & Commitment

### You Are Now a Stock Assessment Practitioner!

::: {.callout-success}
**Next:** Go implement these methods in your fisheries

**Remember:** Good data + simple analysis > complex methods + poor data

**Support:** Network with colleagues, reach out when stuck, share successes
:::

---

## Certificate & Continuation

### Today's Training Included:

- ✓ 3 comprehensive modules
- ✓ Practical exercises
- ✓ Real data analysis
- ✓ R programming skills
- ✓ Decision-making frameworks

### Next Level:
- Advanced modeling (LIME, LBB)
- Ecosystem approaches
- Stakeholder engagement
- Regional coordination

---

## Thank You

**Somali Fish Stock Training**

**Module 3: Data-Limited Assessment Methods**

*Empowering evidence-based fisheries management in Somalia*

---

## Final Activity: Your Commitment

### Reflect and Share:

- What will YOU do first?
- Which stock will you assess?
- How will you share knowledge?
- What support do you need?
- Who will you work with?

---

## Contact & Support

### Training Team:

📧 Email: [training contact]
📱 Phone: [contact number]
🌐 Website: [program website]

### Resources Available:
- Data templates
- R code repository
- Reference documents
- Discussion forum
- Monthly webinars

**Questions? Contact the team anytime!**
