---
title: "Module 1: Stock Assessment Fundamentals & Concepts"
subtitle: "Somali Fish Stock Training"
date: "2025-12-05"
date-modified: today
author: 
    - name: "Dr. Masumbuko Semba"
      affiliation: "Nelson Mandela African Institution of Science and Technology"
      email: "lugosemba@gmail.com"
    - name: "Dr. Rushingisha George"
      affiliation: "Tangania Fisheries Research Institute"
      email: "george.rushingisha@tafiri.go.tz"
    - name: "Dr. Mary Kishe"
      affiliation: "Tanzania Fisheries Research Institute"
      email: "mary.kishe@tafiri.go.tz"
    - name: "Dr. Ismael Kimirei"
      affiliation: "Tanzania Fisheries Research Institute"
      email: "ismael.kimirei@tafiri.go.tz"
format: revealjs
theme: default
transition: slide
background-transition: fade
highlight-style: github
embed-resources: true
---



```{r}
require(sf)
require(tidyverse)
require(sizeMat)
require(gt)
require(tidyplots)
require(patchwork)
require(leaflet)
```

## Welcome to Module 1

**Stock Assessment Fundamentals & Concepts**

:::: {.columns}
::: {.column width="50%"}
### Learning Objectives
- Understand core stock assessment principles
- Identify different data types used in assessments
- Apply concepts to real Somali fisheries data
- Build foundation for advanced methods
:::

::: {.column width="50%"}
### Why This Matters
- Essential for evidence-based management
- Reinforces foundational knowledge
- Prepares for data-limited methods
- Supports sustainable fisheries
:::
::::

---

## What is Stock Assessment?

Stock assessment is the **scientific process** of estimating the status and productivity of fish populations to inform management decisions.

### Key Components:

::: {.incremental}
- **Population Dynamics**: How fish populations change over time
- **Catch Data**: What we remove from the stock
- **Biological Data**: Growth, reproduction, mortality
- **Environmental Context**: Habitat and external factors
:::

---

## The Assessment Cycle

```{r}
#| echo: false
#| fig-align: center

tibble(
  step = c("1. Data\nCollection", "2. Analysis", "3. Stock\nStatus", "4. Advice", "5. Management"),
  x = c(1, 2, 3, 4, 5),
  y = c(1, 1.5, 1, 0.5, 1)
) |>
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 8, color = '#1976D2', alpha = 0.8) +
  geom_path(color = '#666666', size = 1.2, arrow = arrow(type = 'closed', length = unit(0.3, 'cm'))) +
  geom_text(aes(label = step), nudge_y = 0.25, size = 3.5, fontface = 'bold') +
  ylim(0, 2) +
  xlim(0.5, 5.5) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )
```

---

## Types of Data in Stock Assessment

### Primary Data Categories:

::: {.columns}
::: {.column width="50%"}
**Catch Data**
- Total landings (numbers or weight)
- Species composition
- Temporal patterns
- Geographic distribution
:::

::: {.column width="50%"}
**Biological Data**
- Size distributions
- Age/length structure
- Growth rates
- Maturity information
:::
:::

---

## Types of Data (continued)

::: {.columns}
::: {.column width="50%"}
**Effort Data**
- Number of vessels/fishers
- Fishing days
- Landing site activity
- Spatial distribution
:::

::: {.column width="50%"}
**Environmental Data**
- Oceanographic conditions
- Habitat quality
- Seasonal patterns
- Climate factors
:::
:::

---

## Case Study: Somali Fisheries Data

### What We Have in Our Dataset:

::: {.incremental}
- **Landing records**: 3,571 specimens across 6 regions
- **Species groups**: Tuna, Mackerel, Shark, Jacks, Billfish, Rays & Skates
- **Time series**: 2017-2023 (7 years of data)
- **Biological measurements**: Length (cm) and Weight (kg)
- **Geographic coverage**: Berbera, Borama, Mogadishu, Kismayo, Baidoa, Galkacyo
:::

---

## Key Indicators from Our Data

### CPUE (Catch Per Unit Effort)

Specimens caught per landing site per year

:::: {.columns}
::: {.column width="50%"}
**What it tells us:**
- Stock abundance trends
- Fishing pressure changes
- Sustainability signals
:::

::: {.column width="50%"}
**Warning signs:**
- ⚠️ Declining CPUE = potential overfishing
- ✓ Stable CPUE = sustainable harvest
:::
::::

---

## Key Indicators: Size Structure

### Average Fish Size Trends

::: {.incremental}
- **Healthy stocks**: Stable or increasing average size
- **Overfished stocks**: Declining average size over time
- **Young populations**: Smaller specimens dominate
- **Mature populations**: Mix of sizes present
:::

### Why Size Matters:
Smaller fish = fewer breeding adults = recruitment problems

---

## Key Indicators: Species Diversity

### Shannon Diversity Index

Measures ecosystem stability:

::: {.columns}
::: {.column width="50%"}
**High Diversity**
- Many species equally represented
- Ecosystem resilience
- Lower collapse risk
- Adaptable to changes
:::

::: {.column width="50%"}
**Low Diversity**
- Few dominant species
- Ecosystem vulnerability
- High collapse risk
- Susceptible to shifts
:::
::::

---

## Key Indicators: Seasonal Patterns

### Why Seasonality Matters:

::: {.incremental}
- **Peak seasons**: When stocks are most vulnerable
- **Off-season monitoring**: Essential for baseline data
- **Breeding seasons**: Critical conservation periods
- **Nomadic patterns**: Species movement and migration
:::

### Somalia's Monsoon Seasons:
- **NEM (Nov-May)**: Northeast Monsoon
- **SEM (Jun-Sep)**: Southwest Monsoon

---

## Stock Status Assessment Framework

### Traffic Light System:

::: {.columns}
::: {.column width="33%"}
**Green** ✓
- CPUE stable
- Size structure healthy
- High diversity
- Sustainable
:::

::: {.column width="33%"}
**Yellow** ⚠️
- CPUE declining
- Size shrinking
- Diversity moderate
- Caution needed
:::

::: {.column width="33%"}
**Red** ✗
- CPUE low
- Size very small
- Low diversity
- Urgent action
:::
::::

---

## Applying This to Somali Data

### Our Findings Show:

::: {.incremental}
1. **Regional differences**: Some areas more productive than others
2. **Temporal trends**: Changes over 2017-2023 period
3. **Species-specific patterns**: Different groups show different trends
4. **Seasonal variability**: Predictable patterns across years
:::

---

## Data Quality Matters

### Good Data Requires:

::: {.columns}
::: {.column width="50%"}
**Collection Standards**
- Consistent methods
- Regular sampling
- Complete records
- Training & protocols
:::

::: {.column width="50%"}
**Quality Control**
- Data verification
- Error detection
- Outlier assessment
- Metadata documentation
:::
::::

---

## Decision-Making Framework

### From Data to Decisions:

```{r}
#| echo: false
#| fig-align: center

tibble(
  stage = c("Data\nCollection", "Analysis", "Indicators", "Assessment", "Advice"),
  uncertainty = c(20, 35, 25, 40, 50)
) |>
  ggplot(aes(x = factor(stage, levels = stage), y = uncertainty)) +
  geom_col(fill = '#FF6F00', alpha = 0.8) +
  geom_text(aes(label = paste0(uncertainty, "%")), vjust = -0.5, size = 4, fontface = 'bold') +
  labs(
    title = "Uncertainty Increases Through Assessment Chain",
    x = "Assessment Stage",
    y = "Uncertainty Level"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 11, face = 'bold'),
    plot.title = element_text(size = 12, face = 'bold', hjust = 0.5)
  )
```

---

## Key Takeaways

### Module 1 Summary:

::: {.incremental}
1. Stock assessment is systematic, cyclical process
2. Multiple data types needed for robust conclusions
3. Simple indicators reveal important patterns
4. Somali data shows clear regional & temporal trends
5. Quality data essential for confident decisions
6. Uncertainty increases through analysis chain
:::

---

## Next Steps

### Transition to Module 2:

- **Module 1**: ✓ Concepts established
- **Module 2**: Data collection protocols & sampling design
- **Module 3**: Data-limited assessment methods

### Homework:

::: {.callout-note}
Review the landing data summaries from Module 1 analysis. Come prepared to discuss:
- What regional patterns stand out?
- Which species show concerning trends?
- What questions would you ask about the data?
:::

---

## Questions & Discussion

### Let's Explore:

- What surprises you about the data?
- Which indicators seem most important?
- How would you communicate these findings?
- What additional data would help?

---

## References & Resources

### Key Concepts:
- FAO Stock Assessment: https://www.fao.org/fishery/
- R for Fisheries: https://www.afsc.noaa.gov/
- Data-limited Methods: Costello et al., 2012

### Somali Context:
- Somalia Fisheries Profile
- Indian Ocean Tuna Commission (IOTC)
- Regional fisheries management bodies

---

## Thank You

**Somali Fish Stock Training**

Module 1: Fundamentals & Concepts

*Building capacity for evidence-based fisheries management*

Questions? Contact the training team.
