---
title: "Presentation 10: Trophic Level"
subtitle: "Somali Fish Stock Training"
date: "2025-12-15"
author: 
    - name: "Mwijage"
      affiliation: "TAFIRI"
format: revealjs
theme: default
transition: slide
background-transition: fade
highlight-style: github
embed-resources: true
execute:
  eval: true
  echo: true
  warning: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
require(tidyverse)
require(tidyplots)
require(gt)

# Load data
lw <- readxl::read_excel('../data_shared/ProjKalluun_master_data_entry.xlsx', sheet = 'Section 3', skip = 4) |> 
    janitor::clean_names() |> 
    filter(!is.na(landing_site)) |> 
    mutate(
      year = year(date_dd_mm_yy),
      month = month(date_dd_mm_yy, label = TRUE),
      .after = date_dd_mm_yy
    )
```

## Welcome to Presentation 10

**Trophic Level: Understanding the Marine Food Web**

:::: {.columns}
::: {.column width="50%"}
### Learning Objectives
- Define what a trophic level is.
- Understand its importance in fisheries management.
- Learn about "fishing down the food web."
- Calculate the Mean Trophic Index of the catch.
:::

::: {.column width="50%"}
### Why This Matters
- Provides an ecosystem-level view of fishing impacts.
- A key indicator for Ecosystem-Based Fisheries Management (EBFM).
- Helps detect long-term changes in the marine environment.
:::
::::

---

## What is a Trophic Level?

A trophic level is the position an organism occupies in a food chain. It represents a "step" in the flow of energy.

*   **Level 1**: **Primary Producers** (e.g., Phytoplankton) - Create their own food.
*   **Level 2**: **Primary Consumers** (Herbivores, e.g., Zooplankton) - Eat producers.
*   **Level 3**: **Secondary Consumers** (Carnivores, e.g., Mackerel) - Eat primary consumers.
*   **Level 4**: **Tertiary Consumers** (Carnivores, e.g., Tuna) - Eat secondary consumers.
*   **Level 5**: **Apex Predators** (e.g., large Sharks) - Top of the food chain.

---

## The Marine Food Web

In reality, it's a complex web, not a simple chain. A single species can feed at multiple trophic levels.

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
knitr::include_graphics("https://www.researchgate.net/profile/Francisco-Arreguin-Sanchez/publication/237699929/figure/fig1/AS:669456224948248@1536622154429/A-simplified-marine-food-web-showing-the-flow-of-energy-and-biomass-from-lower-to.png")
```

*Source: Arreguín-Sánchez (2014)*

---

## Why is Trophic Level Important?

1.  **Ecosystem Health Indicator**: The average trophic level of the catch can signal major shifts in the ecosystem.
2.  **"Fishing Down the Food Web"**: This is a classic sign of overfishing. Fisheries first deplete large, high-trophic-level predators and then shift to targeting smaller, lower-trophic-level species.
3.  **Management Strategy**: High-trophic-level species (like sharks) are often long-lived and slow to reproduce, requiring more conservative management than low-trophic-level species (like sardines).

---

## Calculating Mean Trophic Index (MTI) of the Catch

We can estimate the health of the fishery by tracking the average trophic level of the fish we catch over time.

**Step 1: Assign Trophic Levels to Fish Groups**

We use known values from sources like FishBase.

```{r}
#| echo: true
trophic_levels <- tibble::tribble(
  ~fish_group,       ~trophic_level,
  "Tuna",            4.4,
  "Shark",           4.5,
  "Billfish",        4.5,
  "Jacks",           3.9,
  "Mackerel",        3.6,
  "Rays and Skates", 3.5,
  "Other",           3.7  # Average assumption
)
```

---

## Calculating MTI of the Catch (cont.)

**Step 2: Calculate the weighted mean trophic level for each year.**

```{r}
#| echo: true
mti_by_year <- lw |>
  filter(!is.na(fish_group)) |>
  left_join(trophic_levels, by = "fish_group") |>
  group_by(year) |>
  summarise(
    mti = mean(trophic_level, na.rm = TRUE),
    .groups = 'drop'
  )
```

**Step 3: Visualize the trend.**

```{r}
mti_by_year |>
  filter(!is.na(year)) |> 
  tidyplot(x = year, y = mti) |>
  add_data_points(size = 4) |>
  add_curve_fit(method = 'loess', linewidth = 1.2) |>
  adjust_y_axis(title = "Mean Trophic Index (MTI) of Catch") |>
  adjust_x_axis(title = "Year", breaks = seq(min(mti_by_year$year, na.rm=T), max(mti_by_year$year, na.rm=T), by = 1)) |>
  adjust_title("Trend in Mean Trophic Level of Somali Catch")
```

**Interpretation**: A consistent downward trend would be a strong warning sign of "fishing down the food web." Our data shows some fluctuation but no clear, steep decline yet.

---

## Key Takeaways

-   Trophic level is a fundamental concept for understanding an organism's role in the ecosystem.
-   Tracking the **Mean Trophic Index (MTI)** of the catch is a powerful, simple indicator of ecosystem health.
-   A declining MTI is a classic sign of **"fishing down the food web,"** indicating unsustainable pressure on top predators.
-   This analysis is a core component of **Ecosystem-Based Fisheries Management (EBFM)**.

---

## Questions & Discussion

**Thank You**
