---
title: "Module 3: Data Collection Methods"
date: "2025-12-05"
date-modified: today
author: 
    - name: "Masumbuko Semba"
      affiliation: "Nelson Mandela African Institution of Science and Technology"
      email: "lugosemba@gmail.com"
    - name: "Dr. Rushingisha George"
      affiliation: "Tangania Fisheries Research Institute"
      email: "george.rushingisha@tafiri.go.tz"
    - name: "Dr. Mary Kishe"
      affiliation: "Tanzania Fisheries Research Institute"
      email: "mary.kishe@tafiri.go.tz"
    - name: "Dr. Ismael Kimirei"
      affiliation: "Tanzania Fisheries Research Institute"
      email: "ismael.kimirei@tafiri.go.tz"
execute: 
  warning: false
  message: false
  comment: ""
---

## Overview

Standardized, high-quality data collection is essential for effective stock assessment. This module covers practical protocols for monitoring fisheries.



```{r}
#| echo: false
#| 
require(sf)
require(tidyverse)
require(sizeMat)
require(gt)
require(tidyplots)
require(patchwork)
require(leaflet)
require(janitor)
```


```{r}
#| echo: false


lw = readxl::read_excel('../data_shared/ProjKalluun_master_data_entry.xlsx', sheet = 'Section 3', skip= 4) |> 
    janitor::clean_names() |> 
    filter(!is.na(landing_site))

lw = lw |> 
 mutate(
    year = year(date_dd_mm_yy),
    month = month(date_dd_mm_yy, label = TRUE),
    day = day(date_dd_mm_yy), 
    .after = date_dd_mm_yy
 ) |> 
    mutate(
    weight_kg = as.numeric(weight_kg),
    length_cm = as.numeric(length_cm)
    )
```


```{r}
#| label: tbl-summary
#| tbl-cap: Fish Landings by Month and Region (Summary)


lw |> 
    janitor::tabyl(month, region) |> 
    janitor::adorn_totals(where = c('col',"row")) |> 
    as_tibble()  |>
    gt() |> 
    # tab_header(
    #     title = "Fish Landings by Month and Region (Summary)",
    #     subtitle = "Crosstab with row and column totals"
    # ) |>
    tab_options(
        table.width = px(600), 
        table.font.size = '12px', 
        heading.title.font.size = '16px',
        heading.subtitle.font.size = '12px',
        heading.subtitle.font.weight = 'bold',
        latex.use_longtable = TRUE, 
        data_row.padding = px(6),
        column_labels.padding = px(8)
    ) |>
    data_color(
        columns = Berbera:Mogadishu,
        rows = 1:12,
        method = 'quantile',
        quantiles = 8,
        colors = scales::col_numeric(
            palette = 'YlOrRd',
            domain = NULL
        )
    ) |>
    opt_stylize(style = 6, color = 'gray')
```

## Learning Objectives

By the end of this module, participants will:

- [ ] Design sampling schemes for landing sites
- [ ] Implement quality assurance procedures
- [ ] Train data collectors
- [ ] Handle common field challenges
- [ ] Coordinate regional monitoring

## Sampling Design Principles

### Key Concepts
- **Population:** Target of inference (all landings in a region)
- **Sampling:** Efficient way to estimate population characteristics
- **Stratification:** Dividing population into meaningful groups
- **Random selection:** Reducing bias in sampling

### Sampling Types
- Systematic sampling (every nth landing)
- Stratified sampling (by vessel type, season, species)
- Purposive sampling (typical vessels)
- Adaptive sampling (responsive to conditions)

## Landing Site Monitoring

### Standard Protocol
1. **Pre-landing observation** - vessel and crew info
2. **Catch landing** - species and weight recording
3. **Length sampling** - take random sample for measurement
4. **Biological sampling** - collect specimens for analysis
5. **Data recording** - digital or paper forms

### Quality Assurance Checks
- Duplicate measurements
- Data validation rules
- Supervisor verification
- Regular audits

## Observer Programs

### Onboard Observation
- Vessel selection
- Observer training
- Data collection at sea
- Safety protocols

### Challenges
- Cost and logistics
- Safety at sea
- Language barriers
- Data transmission

## Data Collection Forms

### Essential Information
- Date, time, location
- Vessel characteristics
- Fishing gear details
- Species and quantities
- Individual fish measurements
- Environmental conditions

## Training Data Collectors

### Key Topics
- Species identification
- Measurement techniques
- Form completion
- Data quality
- Safety procedures

### Certification
- Written exam
- Practical demonstration
- Supervised practice
- Ongoing evaluation

## Regional Coordination

### Standardization
- Common forms across regions
- Consistent protocols
- Regular training
- Quality monitoring

### Communication
- Data flow procedures
- Reporting schedules
- Feedback mechanisms
- Issue resolution

## Practical Activities

- [ ] Practice species identification with specimens
- [ ] Complete data entry exercises
- [ ] Role-play landing site scenarios
- [ ] Develop regional protocols

## Resources

- **Forms:** Standardized data collection templates
- **Guide:** Landing site monitoring manual
- **Tools:** Measurement boards, scales, specimen jars

## Assessment

- Practical identification test
- Form completion exercise
- Protocol design project
- Regional coordination plan

---

**Next Module:** Data Processing and Storage (December 17)
