---
title: "Module 2: Fisheries Data Types"
date: "2025-12-05"
date-modified: today
author: 
    - name: "Masumbuko Semba"
      affiliation: "Nelson Mandela African Institution of Science and Technology"
      email: "lugosemba@gmail.com"
    - name: "Dr. Rushingisha George"
      affiliation: "Tangania Fisheries Research Institute"
      email: "george.rushingisha@tafiri.go.tz"
    - name: "Dr. Mary Kishe"
      affiliation: "Tanzania Fisheries Research Institute"
      email: "mary.kishe@tafiri.go.tz"
    - name: "Dr. Ismael Kimirei"
      affiliation: "Tanzania Fisheries Research Institute"
      email: "ismael.kimirei@tafiri.go.tz"
execute: 
  warning: false
  message: false
  comment: ""
---

## Overview

Fisheries stock assessment relies on diverse data types. This module explores what data is available, how it's collected, and how it's used.



```{r}
#| echo: false
#| 
require(sf)
require(tidyverse)
require(sizeMat)
require(gt)
require(tidyplots)
require(patchwork)
require(leaflet)
```


```{r}
#| echo: false


lw = readxl::read_excel('../data_shared/ProjKalluun_master_data_entry.xlsx', sheet = 'Section 3', skip= 4) |> 
    janitor::clean_names() |> 
    filter(!is.na(landing_site))

lw = lw |> 
 mutate(
    year = year(date_dd_mm_yy),
    month = month(date_dd_mm_yy, label = TRUE),
    day = day(date_dd_mm_yy), 
    .after = date_dd_mm_yy
 ) |> 
    mutate(
    weight_kg = as.numeric(weight_kg),
    length_cm = as.numeric(length_cm)
    )
```

```{r}
lw |> 
  # glimpse()
  select(year, month, region,landing_site, fish_group, length_cm, weight_kg) |>
  sample_n(30)  |> 
  gt() |>
  cols_label(
    year = 'Year',
    month = 'Month',
    region = 'Region',
    landing_site = 'Landing Site',
    fish_group = 'Fish Group',
    length_cm = 'Length (cm)',
    weight_kg = 'Weight (kg)'
  ) |>
    tab_options(
        table.width = px(700), 
        table.font.size = '12px', 
        heading.title.font.size = '16px',
        heading.subtitle.font.size = '14px',
        heading.subtitle.font.weight = 'bold',
        latex.use_longtable = TRUE,
        data_row.padding = px(8),
        column_labels.padding = px(8)
    ) |>
    opt_stylize(style = 6, color = 'gray')
```



## Learning Objectives

By the end of this module, participants will:

- [ ] Classify different fisheries data types
- [ ] Understand advantages and limitations of each data type
- [ ] Explain how different data types serve assessment needs
- [ ] Recognize data quality indicators
- [ ] Assess data availability for Somali stocks

## Data Types Covered

### 1. Catch Data
- Landings by species
- Temporal trends
- Spatial information
- Data sources and quality

### 2. Effort Data
- Fishing effort metrics
- Fleet characteristics
- Seasonality
- Regional variations

### 3. Catch Composition Data
- Species composition
- Size composition
- Discard data
- Bycatch information

### 4. Biological Sampling Data
- Length measurements
- Age data
- Maturity observations
- Condition factors

### 5. Environmental Data
- Sea surface temperature
- Oceanographic conditions
- Seasonal patterns
- Climate impacts

## Somali Fisheries Context

### Typical Data Available
- Landing site monitoring (partial)
- Market surveys (limited)
- Observer programs (developing)
- Research surveys (sparse)

### Gaps and Challenges
- Limited historical data
- Spatial and temporal gaps
- Quality inconsistencies
- Limited biological sampling

## Data Quality Framework

```
Data Quality Assessment:
├── Completeness (% of data collected)
├── Accuracy (measurement error)
├── Consistency (within and across regions)
├── Timeliness (how current is the data)
└── Validity (plausibility of values)
```

## Practical Activities

- [ ] Review actual Somali landing site data
- [ ] Identify data gaps for your region
- [ ] Assess data quality issues
- [ ] Propose data improvements

## Resources

- **Reference Data:** Example datasets from IOTC
- **Standards:** FAO Guidelines on Data Collection
- **Tools:** Excel templates for data entry

## Assessment

- Data type identification quiz
- Quality assessment exercise
- Group discussion on Somali data needs

---

**Next Module:** Data Collection Methods (December 16)
